Comparing Machine Learning algorithms for classifying cognitive states from fMRI data

### 2012-11-09 article/article.tex
More references, initial Application section

In fact, Machine learning algorithms have been employed to learn the patterns of activity associated with some experimental condition. Subsequently, these learned patterns can be used on unseen subjects in order to predict experimental conditions from their brain activity patterns. If these predictions are accurate then it means that the identified patterns hold information about the brain activity associated with some experimental task.

Machine learning techniques have been successfully used for decoding the orientation of a stimulus from fMRI scans \cite{ka:2005decoding}, identifying different patterns of activity for animate versus inanimate objects \cite{kr:2008RSA}, and even reconstructing visual input based on information in the fMRI scans \cite{mi:2008re}. The reason that these machine learning techniques work so well is that patterns of voxel activity can be identified rather than a significant deviation from some baseline. Specifically, information can be present in the pattern of voxels that belong to a certain brain region, even when the average signal from that region is zero (i.e., on average it does not differ from the baseline condition) \cite{to:2012rev}.
The general goal of these machine learning studies is to identify what experimental condition a subject is in at a certain time. The state that a person is in during some trial (e.g., looking at a face) is referred to as \emph{cognitive state}, because the cognitive processes that are associated with being in that state are of interest. Specifically, the aforementioned studies try to infer the cognitive state of a subject from the brain activity that is evoked in a subject by performing that trial. For example, in an experiment where subjects look at a set of pictures (e.g. a star and a plus), they would be in a certain cognitive state that is the result of looking at these pictures. One could then attempt to use machine learning on the fMRI signal that is evoked by the the brain activity associated with these cognitive states, and use this to learn what patterns of activity in the brain are associated with that cognitive state. This \emph{StarPlus} experiment has been performed by researches where, in addition to viewing pictures, subjects also were tasked to read sentences containing some true or false statement about the pictures. These data have then been used to successfully infer the cognitive states of looking at either sentences or pictures within single subjects \cite{mi:2003within}, and between subjects \cite{wa:2003betw}.

In this study we use the same StarPlus data as in the aforementioned experiment, and use machine learning algorithms to decode the cognitive states of subjects when they are looking at pictures or when they are looking at sentences. An interesting difference between many applications of machine learning, and this study is that most machine learning algorithms have the goal of achieving the maximum possible discrimination between conditions or classes. However, for fMRI studies it is important that the pattern of activity that is identified by the algorithm is a pattern associated with the cognitive state. When using non-linear classifiers, it is possible to decode features predictive of the cognitive state, where the brain does not actually respond to these features. For example, when decoding the primary visual cortex with high enough spatial resolution, it would be possible to combine simple features with a non-linear neural network to decode the presence or absence of, for example, faces. However, this information is not actually processed in that part of the brain \cite{to:2012rev}. To avoid this problem it is recommended to only use linear classifiers when decoding cognitive states from fMRI data. The classifiers that we compare in this paper are Gaussian Naive Bayesian (GNB) classifiers, Linear Support Vector Machine (LSVM), and a Multivariate Gaussian (MVG) classifier.\\
\indent The goal of this study is to compare the classification performance of the cognitive states associated with looking at a picture or looking at a sentence between subjects. This objective introduces an important issue, namely, how the data from the fMRI scans can be compared between subjects. There are two main ways this could be achieved. First, groups of voxels that are anatomically defined as belonging to a certain brain region, a so-called region of interest (ROI), can be averaged into a single \emph{supervoxel}, which can be compared between subjects. Second, an anatomically defined talairach coordinate system can be used that maps the voxels of each subject onto an anatomically defined coordinate. The ROI averaging it is a rather crude method, and a lot of information of the individual voxels is lost. However, because the ROI's are relatively easy to identify by an expert they can be compared easily between subjects. By using the Talairach coordinate system the data of each individual voxel can be used~\cite{talairach1988co}. However, because the error of doing this is usually in the order of a centimetre, while the size of the voxels are in the order of millimetres, the coordinates are difficult to generalise. Because we aim to do between-subjects comparisons the generalisation performance is important, thus we will use averaged ROIs in order to compare the classifiers \cite{mi:2004coord}.\\

@@

\section{Application}
\label{sec:application}

What can we do with it now?

- classification problem
- prediction
- cognitive state
\section{Methods}
\label{sec:methods}
General approach consisting of three steps, plus basic data acquisition and preprocessing.
\begin{enumerate}
  \item Feature selection
  \item Training
  \item Testing
\end{enumerate}

>>>>

In fact, Machine learning algorithms have been employed in studies to learn the patterns of activity associated with some experimental condition. Subsequently, these learned patterns can be used on unseen subjects in order to predict experimental conditions from their brain activity patterns. If these predictions are accurate then it means that the identified patterns hold information about the brain activity associated with some experimental task.\\

The general goal of these machine learning studies is to identify what experimental condition a subject is in at a certain time. The state that a person is in during some trial (e.g., looking at a face) is referred to as \emph{cognitive state}, because the cognitive processes that are associated with being in that state are of interest. Specifically, the aforementioned studies try to infer the cognitive state of a subject from the brain activity that is evoked in a subject by performing that trial. For example, in an experiment where subjects look at a set of pictures (e.g. a star and a plus), they would be in a certain cognitive state that is the result of looking at these pictures. One could then attempt to use machine learning on the fMRI signal that is evoked by the the brain activity associated with these cognitive states, and use this to learn what patterns of activity in the brain are associated with that cognitive state. This \emph{StarPlus} experiment has been performed by researches where, in addition to viewing pictures, subjects also were tasked to read sentences containing some true or false statement about the pictures. These data have then been used to successfully infer the cognitive states of looking at either sentences or pictures within single subjects \cite{mi:2003within}, and between subjects \cite{wa:2003betw}.\\

In this study we use the same StarPlus data as in the aforementioned experiment, and use machine learning algorithms to decode the cognitive states of subjects while looking at pictures or while looking at sentences. An interesting difference between many applications of machine learning, and this study is that most machine learning algorithms have the goal of achieving the maximum possible discrimination between conditions or classes. However, for fMRI studies it is important that the pattern of activity that is identified by the algorithm is a pattern associated with the cognitive state. When using non-linear classifiers, it is possible to decode features predictive of the cognitive state, where the brain does not actually respond to these features. For example, when decoding the primary visual cortex with high enough spatial resolution, it would be possible to combine simple features with a non-linear neural network to decode the presence or absence of, for example, faces. However, this information is not actually processed in that part of the brain \cite{to:2012rev}. To avoid this problem it is recommended to only use linear classifiers when decoding cognitive states from fMRI data. The classifiers that we compare in this paper are Gaussian Naive Bayesian (GNB) classifiers, Linear Support Vector Machine (LSVM), and a Multivariate Gaussian (MVG) classifier.\\
\indent The goal of this study is to compare the classification performance of the cognitive states associated with looking at a picture or looking at a sentence between subjects. This objective introduces an important issue, namely, how the data from the fMRI scans can be compared between subjects. There are two main ways this could be achieved. First, groups of voxels that are anatomically defined as belonging to a certain brain region, a so-called region of interest (ROI), can be averaged into a single \emph{supervoxel}, which can be compared between subjects. Second, an anatomically defined talairach coordinate system can be used that maps the voxels of each subject onto an anatomically defined coordinate. The ROI averaging it is a rather crude method, and a lot of information of the individual voxels is lost. However, because the ROI's are relatively easy to identify by an expert they can be compared easily between subjects. By using the Talairach coordinate system the data of each individual voxel can be used~\cite{talairach1988co}. However, because the error of doing this is usually in the order of a centimetre, while the size of the voxels are in the order of millimetres, the coordinates are difficult to generalise. Because we aim to do between-subjects comparisons the generalisation performance is important, thus we will use averaged ROIs in order to compare the classifiers \cite{mi:2004coord}.

@@

\section{Application}
\label{sec:application}
Thus far, machine learning techniques have been successfully applied for decoding the orientation of a stimulus from fMRI scans \cite{ka:2005decoding}, identifying different patterns of activity for animate versus inanimate objects \cite{kr:2008RSA}, and even reconstructing visual input based on information in the fMRI scans \cite{mi:2008re}. The reason that these machine learning techniques work so well is that patterns of voxel activity can be identified rather than a significant deviation from some baseline. Specifically, information can be present in the pattern of voxels that belong to a certain brain region, even when the average signal from that region is zero (i.e., on average it does not differ from the baseline condition) \cite{to:2012rev}.

The studies mentioned before all depend on the question of whether fMRI data carry class
information (pattern discrimination) and exploiting this information to perform classification. However, researchers have indicated at least two other basic questions to which classifiers can be applied, being ``where or when
is class information represented?'' (pattern localization) and ``how is
class information encoded'' or ``how does its encoding relate to known
relationships between stimuli'' (pattern characterization)~\cite{pereira2009machine}. While the main focus of this paper is on the application of machine learning techniques in pattern discrimination, we will briefly discuss these two other applications.

\section{Methods}
\label{sec:methods}
General approach consisting of three steps, plus basic data acquisition and preprocessing.
\begin{enumerate}
  \item Feature selection
  \item Training
  \item Testing
\end{enumerate}
